---
title: "Exercises Day 2"
author: "Richard Paquin Morel, adapted from exercises by Christina Maimone"
date: "`r Sys.Date()`"
output: html_document
params:
    answers: TRUE
---


```{r, echo=FALSE, eval=TRUE}
answers<-params$answers
```

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo=answers, eval=answers,
                      warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE)
```

## Load the data

Load the `gapminder` dataset.

```{asis}
### Answer
```

```{r}
gapminder <- read.csv(here::here("data/gapminder5.csv"), stringsAsFactors=FALSE)
gapminder$country <- as.character(gapminder$country)
gapminder$continent <- as.character(gapminder$continent)
str(gapminder)
```


## If Statement

Use an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. Now do the same for 2012.

Hint: use the `any` function.

```{asis}
### Answer
```

```{r}
if (any(gapminder$year==2002)){
    print('Record(s) for the year 2002 found.')
} else {
    print('Record(s) for the year 2002 not found.')
}

if (any(gapminder$year==2012)){
    print('Record(s) for the year 2002 found.')
} else {
    print('Record(s) for the year 2002 not found.')
}
```


## Loop and If Statements

Write a script that finds the mean life expectancy by country for countries whose population is below the mean for the dataset

```{asis}
### Answer
```

```{r}
gapminder_pop_below_mean <- gapminder[gapminder$pop < mean(gapminder$pop),]
gapminder_lifeExp_country <- aggregate(lifeExp~country, data=gapminder_pop_below_mean, FUN=mean)
for (country in unique(gapminder_lifeExp_country$country)){
    country_life = gapminder_lifeExp_country$lifeExp[gapminder_lifeExp_country$country==country]
    print(paste0('life expectancy of ', country, ' is ', country_life))
}
```
Write a script that loops through the `gapminder` data by continent and prints out whether the mean life expectancy is smaller than 50, between 50 and 70, or greater than 70.

```{asis}
### Answer
```

```{r}
for (c in unique(gapminder$continent)){
    mean_life_exp <- mean(gapminder$lifeExp[gapminder$continent==c])
    if (mean_life_exp < 50){
        print(paste0(c, ' has a life expectancy < 50'))
    } else if (mean_life_exp < 70){
        print(paste0(c, ' has a life expectancy between 50 and 70'))
    } else {
        print(paste0(c, ' has a life expectancy > 70'))
    }
}
```


## Exercise: Write Functions

Create a function that given a data frame will print the name of each column and the class of data it contains. Use the gapminder dataset.  Hint: Use `mode()` or `class()` to get the class of the data in each column. Remember that `names()` or `colnames()` returns the name of the columns in a dataset.

```{asis}
### Answer

Note: Some of these were taken or modified from https://www.r-bloggers.com/functions-exercises/
```

```{r writing a function}

print_name_datatype <- function(dataset) {
    for (col in colnames(dataset)){
    print(paste0(col, ': ', mode(gapminder[[col]])))
    }  
}

print_name_datatype(gapminder)
```

Create a function that given a vector will print the mean and the standard deviation of a **vector**, it will optionally also print the median. Hint: include an argument that takes a boolean (`TRUE`/`FALSE`) operator and then include an `if` statement.

```{r}
### Answer
print_mean_sd <- function(v, median=TRUE){
    print(paste0('mean: ', mean(v)))
    print(paste0('standard deviation: ', sd(v)))
    if (median == TRUE){
        print(paste0('median: ', median(v)))
    }
}
print_mean_sd(gapminder$lifeExp, median=FALSE)
print_mean_sd(gapminder$lifeExp, median=TRUE)

```


## Analyzing the relationship 

Use what you've learned so far to answer the following questions using the `gapminder` dataset. Be sure to include some visualizations!

1. What is the relationship between GDP per capita and life expectancy? Does this relationship change over time? (Hint: Use the natural log of both variables.)

```{r GDP per capita vs LifeExp}
plot(gapminder$gdpPercap~gapminder$lifeExp, 
     main='GDP per capita vs life expentancy',
     ylab='GDP per capita',
     xlab='life expentancy')
l1 <- lm(log(gapminder$gdpPercap)~gapminder$lifeExp)
summary(l1)

viz_yx <- function(y, x, main_param, xlab_param, ylab_param){
    fit <- lm(y ~ x)
    plot(y ~ x, main = main_param,
         xlab = xlab_param, ylab = ylab_param)
    lines(x, predict(fit), col = 'blue')
}

years = unique(gapminder$year)

# for (year_param in years){
#     viz_yx(y=log(gapminder$gdpPercap[gapminder$year==year_param]),
#        x=gapminder$lifeExp[gapminder$year==year_param],
#        main=paste0('log of GDP per capita vs life expentancy in ', year_param),
#        ylab=paste0('log of GDP per capita (log) in ', year_param),
#        xlab=paste0('life expentancy ', year_param)
#     )
# }

i <- 1
for (year_param in years){
    y <- log(gapminder$gdpPercap[gapminder$year==year_param])
    x <- gapminder$lifeExp[gapminder$year==year_param]
    fit <- lm(y ~ x)
    color <- i
    if (i == 1){
        main_param = paste0('log of GDP per capita vs life expentancy over years')
        xlab_param = paste0('log of GDP per capita (log) in ', year_param)
        ylab_param = paste0('life expentancy ', year_param)
        plot(y ~ x, main = main_param, xlab = xlab_param, ylab = ylab_param)
        lines(x, predict(fit), col = color)
        
    } else {
        points(y ~ x, col=color)
        lines(x, predict(fit), col = color)
        
    }
    
    i <- i + 1
    
}



```

2. Does the relationship between GDP per capita and life expectacy vary by continent? Make sure you divide the Americas into North and South America.


```{r how GDP per capital and life expectacy vary by continent}
continents <- unique(gapminder$continent)
continents <- continents[-which(continents=='Americas')]
continents <- c(continents, 'North Americas', 'South Americas')
gapminder_gdp_lifeExp <- gapminder[c('gdpPercap', 'lifeExp', 'continent', 'country')]

# separate North America and South America
gapminder_gdp_lifeExp$continent[gapminder_gdp_lifeExp$country=='United States'] <- 'North Americas'
gapminder_gdp_lifeExp$continent[gapminder_gdp_lifeExp$country=='Canada'] <- 'North Americas'
gapminder_gdp_lifeExp$continent[gapminder_gdp_lifeExp$continent=='Americas'] <- 'South Americas'

for (c in unique(gapminder_gdp_lifeExp$continent)){
    viz_yx(y = log(gapminder_gdp_lifeExp$gdpPercap[gapminder_gdp_lifeExp$continent==c]), 
       x = gapminder_gdp_lifeExp$lifeExp[gapminder_gdp_lifeExp$continent==c],
       main = paste0('log of gdp Per capita vs life expectacy of ', c),
       ylab = 'log of gdp Per capita',
       xlab = 'life expectacy')
}



```
